The benefits of reporting critical effect size values\footnote{The manuscript as been posted as a pre-print and has not yet undergone peer-review.}

### 2025-03-04 paper/paper.Rmd
changes

keywords          : critical effect size values, statistical significance, hypothesis testing, effect sizes 
wordcount         : "4856"

@@

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests [@lakens2021practical; @chow1988significance; @cortina1997logic; @hagen1997praise; @haig2017tests; @miller2017hypothesis; @nickerson2000null; @cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest [@cumming2012statistical; @transue2019apa]. Researchers often publish studies with low power for medium to small effect sizes [@szucs2017empirical], with the main justification for the sample size being resource limitations [@lakens2022sample]. However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes [@altoe2020enhancing]. Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant [@kirk1996practical; @anvari2023not]. The risk is that researchers will heuristically or opportunistically accept all effects as potentially important.  For this reason "researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it" [@anvari2023not].

@@

In this section, we provide guidance and formulas for computing standardized critical effect sizes with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's $d$), correlations (Pearson's $r$), and raw and standardized coefficients in linear models. The same equations can be used to calculate unstandardized effect sizes that can be preferred in some situations (e.g., linear models). These formulas have been incorporated into R functions of the package `criticalESvalue` accessible at: [https://github.com/psicostat/criticalESvalue](https://github.com/psicostat/criticalESvalue), and are elaborated upon in the subsequent section.

@@

Meta-analysis allows to pool information from multiple studies related to a specific research question. The main advantage of meta-analysis is pooling multiple studies to obtain a more precise and powerful estimation of the effect. From a statistical point of view, a meta-analysis can be considered as a weighted linear regression with heterogeneous variances. Similarly to standard linear regression, hypothesis testing is performed using Wald $t$ or $z$ tests [@Viechtbauer2015-tn; @Rubio-Aparicio2020-gd]. For this reason, we can apply the same equations used for the linear regression. The main difference and advantage is that the meta-analysis model usually works directly with standardized effect size measures and regression coefficients are already standardized.

@@

Table \@ref(tab:eqsummary) summarises the equations presented above.

```{r eqsummary, echo=FALSE, tab.cap="Summary of the critical effect size equations."}
library(kableExtra)

@@

In this section, we introduce the aforementioned user-friendly implementation of the mathematical computations as functions of the package `criticalESvalue` in R. Here, we demonstrate its application through two examples: one example of a $t$-test on real data and a computation of the critical effect size values for a correlation from sample size. It is important to note that, in general, depending on the researcher's hypotheses, the package allows for the calculation of either two critical effect size values for two-tailed tests or a single critical effect size value for one-tailed tests, for which the direction must be specified. In the Supplementary online materials additional examples can be found for correlation, $t$-test, paired $t$-test, linear models and regression coefficients.

>>>>

keywords          : critical effect size values, statistical significance, hypothesis testing, effect sizes 
wordcount         : "4922"

@@

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests [@lakens2021practical; @chow1988significance; @cortina1997logic; @hagen1997praise; @haig2017tests; @miller2017hypothesis; @nickerson2000null; @cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest [@cumming2012statistical; APA, -@apa2020publication]. Researchers often publish studies with low power for medium to small effect sizes [@szucs2017empirical], with the main justification for the sample size being resource limitations [@lakens2022sample]. However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes [@altoe2020enhancing]. Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant [@kirk1996practical; @anvari2023not]. The risk is that researchers will heuristically or opportunistically accept all effects as potentially important.  For this reason "researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it" [@anvari2023not, p. 504].

@@

In this section, we provide guidance and formulas for computing standardized critical effect sizes with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's $d$), correlations (Pearson's $r$), and raw and standardized coefficients in linear models. Table \@ref(tab:eqsummary) summarises those equations. The same equations can be used to calculate unstandardized effect sizes that can be preferred in some situations (e.g., linear models). These formulas have been incorporated into R functions of the package `criticalESvalue` accessible at: [https://github.com/psicostat/criticalESvalue](https://github.com/psicostat/criticalESvalue), and are elaborated upon in the subsequent section.

@@

Meta-analysis allows to pool information from multiple studies related to a specific research question. The main advantage of meta-analysis is pooling multiple studies to obtain a more precise and powerful estimation of the effect. From a statistical point of view, a meta-analysis can be considered as a weighted linear regression with heterogeneous variances. Similarly to standard linear regression, hypothesis testing is performed using Wald $t$ or $z$ tests [@Viechtbauer2015-tn; @borenstein2021introduction, p. 338]. For this reason, we can apply the same equations used for the linear regression. The main difference and advantage is that the meta-analysis model usually works directly with standardized effect size measures and regression coefficients are already standardized.

@@

```{r eqsummary, echo=FALSE, tab.cap="Summary of the critical effect size equations, see text for details."}
library(kableExtra)

@@

In this section, we introduce the aforementioned user-friendly implementation of the mathematical computations as functions of the package `criticalESvalue` in R (currently hosted on GitHub). Here, we demonstrate its application through two examples: one example of a $t$-test on real data and a computation of the critical effect size values for a correlation from sample size. It is important to note that, in general, depending on the researcher's hypotheses, the package allows for the calculation of either two critical effect size values for two-tailed tests or a single critical effect size value for one-tailed tests, for which the direction must be specified. In the Supplementary online materials additional examples can be found for correlation, $t$-test, paired $t$-test, linear models and regression coefficients.

### 2025-03-04 paper/paper.tex
changes

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{critical effect size values, statistical significance, hypothesis testing, effect sizes\newline\indent Word count: 4856}
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

@@

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests (Chow, 1988; Cohen, 1994; Cortina \& Dunlap, 1997; Gigerenzer et al., 2004; Hagen, 1997; Haig, 2017; Krueger, 2001; Lakens, 2021; Miller, 2017; Nickerson, 2000). Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest (Cumming et al., 2012; Transue, 2019). Researchers often publish studies with low power for medium to small effect sizes (Szucs \& Ioannidis, 2017), with the main justification for the sample size being resource limitations (Lakens, 2022). However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes (Altoè et al., 2020). Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant (Anvari et al., 2023; Kirk, 1996). The risk is that researchers will heuristically or opportunistically accept all effects as potentially important. For this reason ``researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it'' (Anvari et al., 2023).

@@

In this section, we provide guidance and formulas for computing standardized critical effect sizes with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's \(d\)), correlations (Pearson's \(r\)), and raw and standardized coefficients in linear models. The same equations can be used to calculate unstandardized effect sizes that can be preferred in some situations (e.g., linear models). These formulas have been incorporated into R functions of the package \texttt{criticalESvalue} accessible at: \url{https://github.com/psicostat/criticalESvalue}, and are elaborated upon in the subsequent section.

@@

Meta-analysis allows to pool information from multiple studies related to a specific research question. The main advantage of meta-analysis is pooling multiple studies to obtain a more precise and powerful estimation of the effect. From a statistical point of view, a meta-analysis can be considered as a weighted linear regression with heterogeneous variances. Similarly to standard linear regression, hypothesis testing is performed using Wald \(t\) or \(z\) tests (Rubio-Aparicio et al., 2020; Viechtbauer et al., 2015). For this reason, we can apply the same equations used for the linear regression. The main difference and advantage is that the meta-analysis model usually works directly with standardized effect size measures and regression coefficients are already standardized.

@@

Table \ref{tab:eqsummary} summarises the equations presented above.

@@

\begin{table}[H]
\centering
\caption{\label{tab:eqsummary}Summary of the critical effect size equations.}
\centering
\begin{tabular}[t]{ll}
\toprule
\textbf{Model} & \textbf{Equation}\\
\midrule
One-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s}$\\
\cmidrule{1-2}
 & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p}$\\
\cmidrule{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Paired t-test} & $d_{z_c} = \frac{t_c \times \mbox{SE}_b}{s_D}$\\
\cmidrule{1-2}
Between-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p}$\\
\cmidrule{1-2}
 & $r_c = \frac{t_c}{\sqrt{n - 2 + t^2_c}}$\\
\cmidrule{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Correlation} & $r_{z_c} = \frac{z_c}{\sqrt{n - 3}}$\\
\cmidrule{1-2}
Meta-analysis/Regression & $\beta_c = t_c \times \mbox{SE}_{\beta}$\\
\bottomrule
\end{tabular}
\end{table}

@@

In this section, we introduce the aforementioned user-friendly implementation of the mathematical computations as functions of the package \texttt{criticalESvalue} in R. Here, we demonstrate its application through two examples: one example of a \(t\)-test on real data and a computation of the critical effect size values for a correlation from sample size. It is important to note that, in general, depending on the researcher's hypotheses, the package allows for the calculation of either two critical effect size values for two-tailed tests or a single critical effect size value for one-tailed tests, for which the direction must be specified. In the Supplementary online materials additional examples can be found for correlation, \(t\)-test, paired \(t\)-test, linear models and regression coefficients.

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rubio-Aparicio2020-gd}{}}%
Rubio-Aparicio, M., López-López, J. A., Viechtbauer, W., Marín-Martínez, F., Botella, J., \& Sánchez-Meca, J. (2020). Testing categorical moderators in mixed-effects meta-analysis in the presence of heteroscedasticity. \emph{Journal of Experimental Education}, \emph{88}, 288--310. \url{https://doi.org/10.1080/00220973.2018.1561404}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{critical effect size values, statistical significance, hypothesis testing, effect sizes\newline\indent Word count: 4922}
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

@@

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests (Chow, 1988; Cohen, 1994; Cortina \& Dunlap, 1997; Gigerenzer et al., 2004; Hagen, 1997; Haig, 2017; Krueger, 2001; Lakens, 2021; Miller, 2017; Nickerson, 2000). Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest (Cumming et al., 2012; APA, 2020). Researchers often publish studies with low power for medium to small effect sizes (Szucs \& Ioannidis, 2017), with the main justification for the sample size being resource limitations (Lakens, 2022). However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes (Altoè et al., 2020). Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant (Anvari et al., 2023; Kirk, 1996). The risk is that researchers will heuristically or opportunistically accept all effects as potentially important. For this reason ``researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it'' (Anvari et al., 2023, p. 504).

@@

In this section, we provide guidance and formulas for computing standardized critical effect sizes with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's \(d\)), correlations (Pearson's \(r\)), and raw and standardized coefficients in linear models. Table \ref{tab:eqsummary} summarises those equations. The same equations can be used to calculate unstandardized effect sizes that can be preferred in some situations (e.g., linear models). These formulas have been incorporated into R functions of the package \texttt{criticalESvalue} accessible at: \url{https://github.com/psicostat/criticalESvalue}, and are elaborated upon in the subsequent section.

@@

Meta-analysis allows to pool information from multiple studies related to a specific research question. The main advantage of meta-analysis is pooling multiple studies to obtain a more precise and powerful estimation of the effect. From a statistical point of view, a meta-analysis can be considered as a weighted linear regression with heterogeneous variances. Similarly to standard linear regression, hypothesis testing is performed using Wald \(t\) or \(z\) tests (Borenstein et al., 2021, p. 338; Viechtbauer et al., 2015). For this reason, we can apply the same equations used for the linear regression. The main difference and advantage is that the meta-analysis model usually works directly with standardized effect size measures and regression coefficients are already standardized.

@@

@@

\begin{table}[H]
\centering
\caption{\label{tab:eqsummary}Summary of the critical effect size equations, see text for details.}
\centering
\begin{tabular}[t]{ll}
\toprule
\textbf{Model} & \textbf{Equation}\\
\midrule
One-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s}$\\
\cmidrule{1-2}
 & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p}$\\
\cmidrule{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Paired t-test} & $d_{z_c} = \frac{t_c \times \mbox{SE}_b}{s_D}$\\
\cmidrule{1-2}
Between-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p}$\\
\cmidrule{1-2}
 & $r_c = \frac{t_c}{\sqrt{n - 2 + t^2_c}}$\\
\cmidrule{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Correlation} & $r_{z_c} = \frac{z_c}{\sqrt{n - 3}}$\\
\cmidrule{1-2}
Meta-analysis/Regression & $\beta_c = t_c \times \mbox{SE}_{\beta}$\\
\bottomrule
\end{tabular}
\end{table}

@@

In this section, we introduce the aforementioned user-friendly implementation of the mathematical computations as functions of the package \texttt{criticalESvalue} in R (currently hosted on GitHub). Here, we demonstrate its application through two examples: one example of a \(t\)-test on real data and a computation of the critical effect size values for a correlation from sample size. It is important to note that, in general, depending on the researcher's hypotheses, the package allows for the calculation of either two critical effect size values for two-tailed tests or a single critical effect size value for one-tailed tests, for which the direction must be specified. In the Supplementary online materials additional examples can be found for correlation, \(t\)-test, paired \(t\)-test, linear models and regression coefficients.

@@

\leavevmode\vadjust pre{\hypertarget{ref-apa2020publication}{}}%
American Psychological Association. (2020). \emph{Publication manual of the american psychological association} (7th ed.). American Psychological Association. \url{https://doi.org/10.1037/0000165-000}

@@

\leavevmode\vadjust pre{\hypertarget{ref-borenstein2021introduction}{}}%
Borenstein, M., Hedges, L. V., Higgins, J. P., \& Rothstein, H. R. (2021). \emph{Introduction to meta-analysis}. John wiley \& sons.

@@
